Transcriptional antitermination integrates the expression of loci of diverse phage origin in the chimeric Bartonella Gene Transfer Agent BaGTA

Version 1

Abstract

Abstract

Gene Transfer Agents (GTAs) are mobile genetic elements derived from bacteriophages that mediate genome-wide horizontal gene transfer (HGT) in diverse groups of prokaryotes. BaGTA, encoded by all the pathogens of the genus Bartonella, is a chimeric GTA that evolved by the domestication of two phages. The run-off-replication module ROR of one phage is integrated with the capsid production, DNA packaging and lysis machinery Bgt of a second phage. Restricted to a self-sacrifying subset of the bacterial population, the position-specific DNA amplification and packaging of a genomic plasticity region enriched for genes involved in host interaction and adaptation selectively enhances the HGT frequency of these pathogenicity genes. This feature of BaGTA is considered a key innovation underlying the evolutionary success of Bartonella. Little is known, however, about the mechanism mediating the coordinated expression of the ror and bgt loci. Here, we established the regulatory hierarchy, with ror acting upstream of the capsid gene cluster bgtA-K. BrrG, encoded by the ror locus, controls the transcription of the bgtA-K operon by functioning as a processive antiterminator. This study provides the first insights into the mechanism controlling the coordinated expression of the two BaGTA modules of divergent phage origin. Beyond BaGTA, we propose that antitermination is a broadly relevant mechanism for controlling HGT by GTAs of the Alphaproteobacteria.

Introduction

Timed and tightly controlled gene expression is crucial in bacterial cells. A variety of mechanisms to control transcription are known, comprising both positive and negative regulators (reviewed in Turnbough, 2019). One of the mechanisms of negative regulation involves transcription attenuation at termination sites (reviewed in Santangelo and Artsimovitch, 2011). The regulation of transcription attenuation is achieved by a diverse class of antiterminators which counteract the attenuation (Santangelo and Artsimovitch, 2011). Antiterminators fall into two categories and can either act specifically at a single, defined terminator site or convert RNA polymerase into a termination-resistant complex that is generally resistant to multiple termination sites over a longer distance (Santangelo and Artsimovitch, 2011; reviewed in Goodson and Winkler, 2018). In the latter case, such antiterminators are referred to as processive antiterminators. Well-known examples of processive antiterminators are Q and N of bacteriophage lambda which control progression of the infection cycle (Goodson and Winkler, 2018).

Peculiar examples of phage evolution are Gene Transfer Agents (GTAs) (reviewed in Québatte and Dehio, 2019; Banks and Tung, 2024; Craske et al., 2024). GTAs are mobile genetic elements, which emerged from defective temperate bacteriophages. GTAs encode bacteriophage-like particles and in contrast to phages transfer random pieces of host bacterial DNA. Despite their phage origin, GTAs are not self-transmitting because the DNA packaging capacity of a GTA particle is too small to package the entire gene cluster encoding it. Therefore, GTAs are transferred vertically (Banks and Tung, 2024; Craske et al., 2024; Kogay et al., 2022; Lang et al., 2012). GTAs are found both in bacteria and archaea, and are one of the non-canonical mechanisms of horizontal gene transfer (HGT) (Arnold et al., 2022).

Bacterial pathogens of the genus Bartonella encode a chimeric GTA known as BaGTA (Québatte and Dehio, 2019; Berglund et al., 2009; Guy et al., 2013). BaGTA encodes the gene clusters bgtA-K, responsible for the synthesis and assembly of GTA particles, bgtL-R, with orphan functions, and bgtS-V, involved in host cell lysis (Fig. 1A). DNA packaging in GTA particles is coupled to a process of site-specific DNA amplification mediated by the ror locus located in between the bgtL-R and bgtS-V gene clusters. This locus, emerging from a different phage domestication event, includes an origin of run-off replication (ROR) and the associated brr gene cluster (brrA-G) (Berglund et al., 2009; Guy et al., 2013; Québatte et al., 2017; Tamarit et al., 2018) (Fig. 1A). The bgt and ror loci are located within a high plasticity zone of the genome that is prone to DNA integrations and is highly enriched for genes encoding virulence properties involved in host interaction and adaptation (Berglund et al., 2009; Guy et al., 2013). This specific genome architecture leads to ROR-driven amplification and packaging of these pathogenicity genes, resulting in elevated transfer rates to other Bartonella bacteria compared to housekeeping genes encoded more distantly to the ROR origin (Québatte et al., 2017). This process is considered to provide the recombination substrates for the duplication and evolutionary diversification of genes encoding the type IV secretion systems (T4SSs) and cognate effector proteins which collectively represent a key innovation associated with the adaptive radiation of these pathogens to diverse mammalian hosts (Guy et al., 2013; Engel et al., 2011). Additionally, it is also considered to facilitate the fast evolution of surface-located autotransporters serving in host interaction and adaptation, such as the recombination-driven antigenic variation of the CFA autotransporter (Siewert et al., 2022). Thus, BaGTA is a crucial factor for the fitness and adaptability of pathogens in the Bartonella genus (Québatte and Dehio, 2019). However, little is known about how the expression of the bgt and brr loci derived from separate phages is regulated in a coordinated manner.

BrrG encodes a putative antiterminator essential for bgtA-K expression and BaGTA-mediated gene transfer.

(A) Schematic representation of the BaGTA locus. Upper part: Overview of the bgt and ror gene clusters. Lower part: Close-up of the bgtA-K locus encoding capsid and packaging functions and of the ror locus, encoding the run-on-replication origin oriROR and the associated brrA-G gene cluster. For the reporter strain bgtC-gfp and its derivatives used in D-G, the insertion site of the transcriptional gfp reporter within the gtaA-K locus is indicated. (B) Comparison of the structure of 21Q antiterminator (6p1b) of phage P21 and Alphafold2 model of BrrG (A0A0R4J952). Similar structural features are presented in identical colour. (C) Schematic presentation of (i) the flow cytometry assay to monitor expression of the bgtA-K locus using the bgtC-gfp transcriptional reporter strain (upper panel) and (ii) the assay to monitor GTA transfer via transfer of GmR from the GTA-positive donor strain to the ΔbgtCD SpR GTA-negative (ΔGTA) recipient strain resulting in SpRGmR bacteria and formula to calculate the frequency of double resistant bacteria (fDRB) as a proxy for GTA transfer. (D) BrrG overproduction increases the bacterial population expressing bgtA-K. BrrG expression was induced in strain bgtC-gfp (pPtet-brrG) with anhydrotetracycline (ATc). Strain bgtC-gfp H1GmR (bgtC-gfp BH01490::Himar1GmR) displaying wild-type properties served as a positive control. (E) BrrG overproduction does not enhance GTA transfer. Strain bgtC-gfp H1GmR displaying wild-type properties and the GTA-negative mutant ΔbgtCD H1GmR (ΔbgtCD BHRS00695::Himar1GmR) served as a positive and negative control, respectively. (F) Deletion of brrG (ΔbrrG bgtC-gfp H1GmR [ΔbrrG BH01490::Himar1GmR]) results in reduced expression of the bgtA-K locus. (G) Overproduction of BrrG in a ΔbrrG mutant background restores GTA transfer frequency to the wild-type level. BrrG expression was induced in strain ΔbrrG H1GmR (pPtet-brrG) with ATc. (D-G) The time-course experiments were started with the incubation of CBA agar-grown bacteria in M199 medium supplemented with 10% FCS. 25 ng/ml Atc was added where indicated to induce expression by the Ptetpromoter. The experiments were performed in biological triplicates (D-F) or quadruplicates (G), error bars indicate SD.

In this study, we elucidated the regulatory hierarchy between the bgtA-K and ror loci, resolving the mechanism of gene expression control. We discovered that BaGTA operates under a strict regulatory framework, with a protein encoded by the ror locus and controlling the expression of the bgtA-K capsid-production locus. Specifically, our research revealed that in absence of BaGTA activation, transcription of the bgtA-K locus is attenuated by a terminator site. Upon BaGTA activation by a yet unknown process, the ror locus-encoded BrrG overcomes this transcription attenuation of the bgtA-K locus by acting as a processive antiterminator. Furthermore, we propose that antitermination is a common mechanism of GTA regulation in the Alphaproteobacteria. Collectively, these findings offer significant insights into the regulation of BaGTA-mediated gene transfer in pathogens of the genus Bartonella and suggest a broader impact of processive antitermination as a regulatory principle in related GTAs.

Results

BrrG shows structural similarity to the P21 phage-encoded Q21 antiterminator and is essential for GTA transfer and expression of capsid components

Previously, we identified by multiplexed transposon sequencing the genes of BaGTA encoding both for donor transfer and recipient uptake functions (Québatte et al., 2017). However, many of the identified genes had only hypothetical or no functional sequence-based annotation, leaving their contribution to BaGTA function unclear. To address this knowledge gap, we employed the protein structure prediction tool AlphaFold2 to model unannotated proteins of BaGTA genes and searched the Protein Data Bank (PDB) database for similar structures. We discovered that the brrG gene (Fig. 1A), which is located in the ror locus, encodes a protein with structural similarity to the antiterminator Q21 of the lambdoid bacteriophage P21 (Fig. 1B). Q21 and its counterpart Q from phage λ (Qλ) function as processive antiterminators involved in bypassing intrinsic terminators (Guo et al., 1991; Grayhack and Roberts, 1982). This enables uninterrupted expression of the late genes, which encode capsid morphogenesis and lysis cassettes. Therefore, we hypothesized that, similar to Q21 and Qλ, BrrG represents a putative antiterminator controling gene expression within the BaGTA regulatory network.

To test this hypothesis, we assessed the effect of brrG overexpression on the transcription of the bgtA-K locus that encodes components of the capsid synthesis and assembly machinery as essential components of BaGTA production. To address this, we utilized the previously established reporter strain bgtC-gfp that contains a promoterless gfpmut2 inserted in the bgtA-K locus in the intergenic region between bgtC and bgtB (Québatte et al., 2017) (Fig. 1A). This transcriptional reporter enabled us to monitor bgtA-K expression over time using flow cytometry (Fig. 1C, upper panel). We observed that plasmid-based overproduction of BrrG increased the fraction of GFP fluorescent bacteria to up to 75%, compared to about 15% at native BrrG expression (Fig. 1D), thus massively expanding the fraction of cells expressing the bgtA-K locus. To test whether this increase is followed by GTA production and elevated DNA transfer, we performed the GTA transfer assay. For that, we cocultivated donor and recipient cells carrying different resistance markers followed by plating on selective media to enumerate double-resistant bacteria as a read-out for the frequency of GTA-mediated gene transfer (Fig. 1C, lower panel). We observed that - while the proportion of cells expressing the BaGTA locus increased as a result of BrrG overproduction, GTA transfer rates remained similar to native BrrG expression (Fig. 1E). In contrast, deleting brrG led to a significant decrease in the proportion of fluorescent bacteria (Fig. 1F) and caused DNA transfer rates to drop at least 100-fold compared to those observed at native BrrG expression (Fig. 1G). Together, these findings indicate that BrrG plays a crucial role in regulating BaGTA. It is capable of activating the expression of capsid components in a large part of the population that is initially GTA inactive. However, overproduction of BrrG is not sufficient to increase the rate of DNA transfer, indicating that it acts in concert with other regulators controling other crucial steps of this complex process.

BrrG binds in proximity to the PbgtA-K promoter located upstream of to the putative bgtA-K operon

In lambdoid phages, both Q21 and Qλ are DNA-binding proteins that recognize a motif located in the promoter region of the late operon, which encodes lysis cassettes and proteins required for capsid morphogenesis (Guo et al., 1991; Grayhack and Roberts, 1982; Yin et al., 2019; Shi et al., 2019; Yin et al., 2022). Considering the structural similarity between BrrG and Q21, we hypothesised that BrrG may be functionally conserved and thus able to directly bind to DNA at the yet unidentified promoter of the bgtA-K locus. To test this, we performed chromatin immunoprecipitation of epitope-tagged BrrG followed by DNA sequencing. We found that BrrG binds to DNA, with a single binding site located in the intergenic region far upstream of the bgtK gene representing the first conserved gene of the bgtA-K locus (Fig. 2A) (Guy et al., 2013; Tamarit et al., 2018). Given this remote upstream location of the BrrG binding site, we wondered whether the promoter elements might also be located nearby. First, we tested whether the 999 bp fragment spanning the region upstream of bgtK until the end of clpB could drive expression of a fused promoterless dsRed gene using flow cytometry as readout (Fig. 2A, plasmid pPf3-dsRed). Strikingly, this fragment did not only demonstrate promoter activity but also displayed at the population level a temporal control resembling closely the expression pattern of the chromosomal transcriptional reporter strain bgtC-gfp H1GmR with a characteristic peak in the size of expressing population at the 6 h time-point of the 24 h expression kinetic (Fig. 2B). This finding suggests that the bgtA-K locus is organized as a transcriptional unit (operon) that is driven by a promoter localizing to the 999 bp fragment upstream of bgtK, which we thus named PbgtA-K (Fig. 2A).

BrrG binds in proximity to the PbgtA-K promoter located upstream of the putative bgtA-K operon.

(A) Location of the PbgtA-K promoter and of the BrrG binding site identified by ChIP-Seq (gray histogram). The fragments upstream of the bgtA-K locus tested for promoter activity in promoter-probe plasmids as indicated. (B) The fusion of the 999 bp fragment spanning from the region between bgtK and clpB to dsRed (strain WT (PPf3-dsRed) displays a temporal transcriptional profile similar to the chromosomal transcription reporter for the bgtA-K locus (strain bgtC-gfp H1GmR [bgtC-gfp BH01490::Himar1GmR]). (C) The PbgtA-K promoter is located far upstream of the bgtA-K locus. Left: Temporal transcriptional profiles controlled by fragments spanning nucleotide 1-416 (plasmid psF1), nucleotides 271-782 (plasmid psF2), or nucleotides 594-999 (plasmid psF3) of the 999 bp fragment in Pf3. Right: Corresponding histograms of florescence distribution at the 6 h time-point. Strain bgtC-gfp displaying wild-type expression pattern of the bgtA-K locus was used as control. (D) DNA sequence alignment of the predicted promoter region within psF3 fragment with homologous sequences from various Bartonella species. The predicted -35 and -10 promoter elements and the inferred transcriptional start site are indicated. Numbers in parentheses indicate the location of the promoter elements relative to bgtK. (B, C) The time-course experiments were started with the incubation of CBA agar-grown bacteria in M199 medium supplemented with 10% FCS. The experiments were done in biological triplicates, error bars indicate SD.

Next, to narrow down the location of PbgtA-K, we split the 999 bp fragment into three partially overlapping fragments and tested them with the same approach using a gfpmut2-based promoter-probe plasmid (plasmids psF1, psF2 and psF3, Fig. 2A). We observed that the most upstream fragment spanning nucleotides 594-999 was able to drive broad expression of gfpmut2 at population level (plasmid psF3 in Figure 2C). The fraction of fluorescent bacteria was substantially higher compared to the chromosomal transcriptional reporter strain bgtC-gfp, expanding beyond 40% at each time-point (Fig. 2C). Importantly, the lack of temporal regulation as seen by the chromosomal transcriptional reporter strain bgtC-gfp and likewise the 999 bp fragment in plasmid Pf3 (compare Figs. 2B and 2C) suggests that absence of downstream control elements located in the more proximal region spanning nucleotides 1-593 converted PbgtA-K into a constitutive state. To map PbgtA-K based on conserved elements (-10 and -35 boxes) we aligned corresponding sequences from a selection of Bartonella species covering the genomic diversity of this genus. By this comparative approach, we identified a highly conserved sequence motive (TTGACA-N17-TATCAT) that shows almost full identity to the canonical sigma-70 promoter sequence of E. coli (Fig. 2D) (Campbell et al., 2002; Murakami et al., 2002). This suggests that PbgtA-K functions as a minimal promoter of the bgtA-K locus. Interestingly, the BrrG binding region coincides with PbgtA-K, suggesting that BrrG controls transcription either by direct binding to PbgtA-K or in the vicinity of it.

Taken together, we identified PbgtA-K as a strong constitutive promoter located far upstream of the bgtA-K locus, driving its expression as a transcriptional unit along with more proximal elements required for controlling temporal expression at the population level. Additionally, we found that the putative antiterminator BrrG, similar to Q21 and λQ, is a DNA-binding protein that directly binds to PbgtA-Kand likely controls bgtA-K operon expression at population level.

A long leader sequence attenuating expression is located downstream of the PbgtA-K promoter

To confirm that the identified conserved sequence is the promoter and to hint at the function of the long downstream sequence in transcription, we generated a panel of promoter-probe plasmids harboring truncations of the 999 bp fragment bearing promoter activity (pF0-gfp to pF6-gfp) (Fig. 3A). We observed that the 29 bp fragment spanning the predicted PbgtA-K displayed fluorescence in the vast majority of cells, confirming the promoter prediction (F0 fragment, Fig. 3B). Similar promoter activities were recorded for the fragments F1 and F2 that are extended by 27 bp or 133 bp toward the bgtA-K locus, respectively. A striking difference in fluorescence distribution with an almost full loss of fluorescent bacteria was observed for the F3 fragment containing, relative to F2, an additional 139 bp (Fig. 3B). Fragments F4-F6 with further downstream extensions also displayed pattern of low fluorescence, while a bi-model distribution with low- and intermediate-expressing bacteria was observed for fragments F4 and F6.

A putative terminator is encoded in the long non-protein coding sequence downstream of the PbgtA-K promoter encodes.

(A) Schematic representation of the 5’-part of the bgtA-K locus with the truncations tested with promoter-probe plasmids. Locations of the PbgtA-Kpromoter and the predicted terminator are shown. (B) Histograms of fluorescence distribution measured for the bacterial populations containing the promoter-probe plasmids with various truncations at 6 h. (C) Predicted secondary structure and free energy of the putative TbgtA-K terminator. Numbers in parentheses indicate the location of the terminator relative to bgtK. The time-course experiment was started with the incubation of CBA plate-grown bacteria in M199 medium supplemented with 10% FCS. Experiments were performed in biological triplicates, error bars indicate SD.

Because fragment F3 is the shortest fragment displaying reduced expression, we hypothesized that a terminator could be present in the sequence extending the highly fluorescent F2 fragment. Indeed, a putative L-shaped terminator with a free energy (ΔG) of -6 kcal/mol is predicted within this stretch of 139 bp of the F3 fragment, which might explain the observed difference in expression levels (Fig. 3C). Altogether, these results demonstrate that the minimal promoter enables uniform and constitutive expression of the BaGTA locus at population level. The observed temporal regulation and heterogeneity of the gene cluster expression appear to be provided by the stretch of 864 bp located downstream of the promoter and upstream to bgtK as first gene of the bgtA-G locus. This stretch of DNA contains the putative terminator and could harbor other terminators or additional regulatory elements.

A -10-like element is required for expression of the bgtA-K locus

In phage lambda, the process of Q-mediated antitermination relies on the engagement of Q into the complex with RNA-polymerase (RNAP) and the conversion of the latter into a termination-resistant form (Perdue and Roberts, 2011). This process begins with RNAP pausing at the -10- like element, a six-nucleotide motif (NANNNT), located just downstream of the -10 promoter element, which allows Q to engage with the transcription complex (Perdue and Roberts, 2011).

Therefore, we set out to determine whether a -10-like element is present in the vicinity of the identified PbgtA-K minimal promoter. We examined the previously generated alignment and identified a conserved NANNNT site located 5 bp downstream of the identified promoter PbgtA-K(Fig. 4A). As the conserved nucleotides A and T at positions +2 and +6 in phage lambda and other phages are also conserved in this motive directly downstream of PbgtA-K, we hypothesized that this site could act as a -10-like element (Fig. 4A). To validate this hypothesis, we used the promoter-probe plasmid containing the F6 fragment (Fig. 3A), and introduced two point mutations, replacing the conserved A in position -2 to G and T in position +6 to C (pF6(-10mut)-gfp), respectively. Additionally, we introduced a second plasmid directing ATc-inducible BrrG overproduction to test whether these mutations would affect the activity of the putative antiterminator in the flow cytometry assay (Fig. 1C). In contrast to plasmid pF6-gfp encoding the native -10-like element (Fig. 4B, left, -ATc), pF6(-10mut)-gfp did not display the characteristic kinetics of GFP-expression of wild-type bacteria with a peak at 6 h (Fig. 4B, right, -ATc). Moreover, neither did it display the massive expansion of the GFP-expressing population upon BrrG overproduction (+ATc) (Fig. 4B). Thus, the replacement of two conserved nucleotides at the putative -10-like element completely abolished transcription of the downstream bgtA-K locus under both native BrrG expression and BrrG overproduction conditions. This suggests that the identified conserved site is a functional - 10-like element essential for the BrrG-mediated antitermination.

The -10-like element downstream of the PbgtA-K promoter is crucial for the antitermination by BrrG.

(A) Alignment of Bartonella spp. consensus and lambdoid phage sequences (P21, phi80 and lambda) sequences containing PbgtA-Kand the PR’ promoters followed by the -10-like elements, respectively. Promoter -35 and -10 elements, as well as the -10-like elements are marked. (B) Histograms of fluorescence distribution for the bacterial populations containing the promoter-probe plasmid pF6 (compare to Fig. 3A) with the PbgtA-Kpromoter and the adjacent native -10-like element or its derivative, plasmid pF6(-10mut), carrying mutations in two conserved positions of the -10-like element. Plasmid pPtet-brrG encodes an ATc-inducible brrG expression cassette. BrrG overproduction was induced by 25 ng/ml ATc. The time-course experiment was started with the incubation of CBA plate-grown bacteria in M199 medium supplemented with 10% FCS. Bacteria were cultivated in M199 medium supplemented with 10% FCS and ATc inducer at indicated concentrations. Experiments were performed in biological triplicates, error bars indicate SD.

BrrG is a processive antiterminator allowing RNAP to bypass the termination site

We showed that BrrG shares structural features with the processive antiterminator Q21 and that its function is dependent on the -10-like element suggesting that BrrG belongs to the class of processive antiterminators. To validate this hypothesis, we tested the antitermination properties of BrrG towards a synthetic terminator. For this, we used the artificially designed strong terminator BBa_B1006 from the iGEM collection and placed it in between the F2 fragment (see Fig. 3A), which contains the native PbgtA-K promoter and downstream -10-like element, but not the native terminator and the gfp reporter gene (Fig. 5A). The plasmid pF2-gfp containing the terminator-less fragment F2 drives GFP-expression in a large part of the bacterial population at both BrrG native expression and overproduction conditions (Figs. 3B and 5B). In contrast, the derivative pF2-TBBa_1006-gfp containing in addition the artificial terminator showed a strongly reduced population expression of GFP, but a marked increase in GFP-expressing bacteria upon BrrG overproduction (Fig. 5B, right panel). Together, these findings demonstrate that the presence of BrrG allows the RNAP to bypass both the native termination site, as well as an unrelated synthetic terminator, indicating that BrrG is indeed a processive antiterminator.

Antiterminator BrrG allows RNA polymerase to bypass an engineered termination site.

(A) Schematic representation of the experimental setup. Plasmid pF2-TBBa_1006-gfp carries the insertion of the artificial strong terminator TBBa_1006in between the terminatorless promoter fragment F2 and gfp of plasmid pF2-gfp. (B) Histograms of fluorescence distribution for the bacterial populations containing the fragment F2, the fragment F2 with the terminator and the sequence with the mutated -10-like element (pF6-10mut). Time of the measurements is indicated. Induction was done with 25ng/ml ATc. Bacteria were cultivated in M199 medium supplemented with 10% FCS and ATc inducer at indicated concentrations. Experiments were performed in biological triplicates, error bars indicate SD.

Discussion

Processive antitermination is an important mechanism of transcriptional control identified both for bacteria and phages. In temperate phages processive antiterminators, i.e phage lambda N and Q-antiterminators, switch on expression of early and late genes, respectively, and thus enable the progression of the infection cycle (Goodson and Winkler, 2018).

Given the ancestral relationship between GTAs and phages, this study investigates whether transcription attenuation and antitermination are preserved in the chimeric BaGTA and whether they play a role in the regulatory integration of two functional gene clusters of divergent phage origin. We showed that the “late” capsid producing bgtA-K genes belong to a single transcriptional unit derived from one phage, and that their expression is enabled by the antiterminator BrrG, encoded within the ror locus, which originates from a second phage. However, the question remains: Is BrrG a Q-like antiterminator, and does it function in a similar manner?

The canonical Q-dependent regulatory circuit consists of several critical components: Q antiterminator, the promoter PR’, a termination site and a -10-like element (Perdue and Roberts, 2011; reviewed in Casjens and Hendrix, 2015). The first component is the Q antiterminator which binds to DNA and engages into the complex with RNAP. The Q-RNAP complex is termination-resistant and permits RNAP to bypass the termination signals in front of late genes which results in the synthesis of the capsid components (Guo et al., 1991; Grayhack and Roberts, 1982). In our study we showed that BrrG has structural similarity to the Q21 antiterminator of the P21 lambdoid phage. Moreover, BrrG is able to counteract transcription attenuation imposed by an introduced artificial termination site. Additionally, we demonstrated that BrrG binds in proximity to the PbgtA-K promoter and upregulates transcription of the bgtA-K cluster – BaGTA’s counterpart of the late gene cluster.

The second component of the canonical Q-dependent regulatory circuit is the promoter PR’ containing Q-binding site for interaction with Q and engagement with RNAP. In this study, we demonstrated that the PbgtA-K promoter overlaps with the BrrG binding site. Given that the promoter PbgtA-K is located 864 bp upstream from the first gene of the bgtA-K cassette and is separated from the former by a gene BH14100, this raises the question of whether BH14100 belongs to the bgtA-K cluster and what functional role it may play. Previous studies have shown that BH14100 is not conserved among bartonellae and thus is likely not involved in the BaGTA regulation or synthesis (Guy et al., 2013; Tamarit et al., 2018). In support of this, our previous study using multiplexed transposon sequencing showed no effect of BH14100 disruption on GTA transfer, indicating that BH14100 has either no or only a very modest effect on GTA synthesis (Québatte et al., 2017).

The third component of the canonical Q-dependent regulatory circuit is the termination site which attenuates expression of the late gene cluster’. We showed that the putative terminator TbgtA-K is located downstream of the PbgtA-K promoter and regulates expression of the bgtA-K operon. Whereas the bgtA-K locus is only active in the GTA producing cells, deleting the putative terminator TbgtA-K renders broad and constitituve expression of the bgtA-K locus which suggests that the temporal regulation and heterogeneity in expression are granted by TbgtA-K.

The fourth component of the canonical Q-dependent regulatory circuit is the -10-like element located in proximity to the promoter PR’ driving expression of late genes. This element resembles the -10-element of the promoter and defines the sigma-dependent pausing of RNAP which is crucial for the engagement of Q into the complex with RNAP (Perdue and Roberts, 2011). We found that expression of the bgtA-K locus is dependent on the conserved -10-like element located downstream of the PbgtA-K promoter. Altogether, these data suggest that BrrG is an antiterminator found in BaGTA and involved in regulating expression of the bgtA-K cluster. Based on these lines of evidence we propose BrrG as Q-like processive antiterminator.

However, why does GTA-mediated DNA transfer remain unaffected upon brrG overexpression?

A simple explanation could be that bacterial lysis and particle release are controlled independently of BrrG which activates only transcription of the bgtA-K locus. This notion is supported by the apparent absence of lysis of fluorescent bacteria upon BrrG overproduction (Fig. 1D). For the archetypical RcGTA of Rhodobacter capsulatus, particle synthesis and release are controlled differentially by the CckA-ChpT-CtrA phosphorelay system, where the phosphorylation state of CtrA controls the progression from the GTA synthesis to a lytic release (Farrera-Calderon et al., 2021; Fogg, 2019; Westbye et al., 2018; Westbye et al., 2013). We hypothesize that similarly to RcGTA a trigger could be involved in the regulation of bacterial lysis and BaGTA release. BaGTA contains two lysis cassettes – bgtAB and bgtTUV – and only the former one is located within the larger bgt cluster, suggesting that the latter cassette is controlled independently.

Taken together, we demonstrate that the transcriptional control mechanism involving transcription attenuation and antitermination, as described for phage lambda, is functionally conserved in the chimeric BaGTA. This mechanism specifically coordinates the temporal expression of the ROR and bgtA-K loci, which have independent phage origins. The ROR-encoded BrrG acts as a processive antiterminator, facilitating the induction of the bgtA-K transcriptional unit. However, is this regulatory mechanism also conserved in other types of GTA beyond BaGTA of Bartonella? The recent publication by Tran and Le has shown that termination-antitermination is involved in control of gene expression in evolutionary distinct CcGTA of Caulobacter crescentus (Tran and Le, 2024). The authors showed that transcription of the structural genes, which are counterparts of the bgtA-K locus, is co-activated by GafY and IHF, while the GafZ antiterminator prevents expression attenuation at the termination sites. Furthermore, GafYZ regulators of CcGTA are homologous to N’- and C’-domains of GafA – the direct activator of RcGTA (Fogg, 2019; Gozzi et al., 2022; Sherlock and Fogg, 2022), which indicates that RcGTA might also be controlled by processive antitermination. Previous studies of GafA indicated that its C’-domain has similarity to the region 4 of a sigma factor (R4σ) and therefore authors proposed GafA as an alternative sigma factor or a transcription factor (Fogg, 2019; Sherlock and Fogg, 2022). However, the related R4σ-like fold was also found in the resolved structure of Q21 antiterminator indicating that findings of Fogg and Sherlock might, in fact, speak for the implication of GafA in antitermination (Shi et al., 2019). Interestingly, Tran and Le found neither sequence nor structural similarity of GafZ to known antiterminators Qλ and Q21. Our comparison of resolved structure of Q21 and modelled BrrG, GafZ and GafA, clearly indicates similar structural features suggesting both GafZ and GafA being antiterminators (Suppl. Figure 1).

RcGTA and CcGTA, on one hand, and BaGTA on the other, belong to two evolutionary distinct groups of GTA that emerged from independent events of phage domestication (Kogay et al., 2022; Tamarit et al., 2018). The reliance of these groups on the termination-antitermination mechanism of transcription control clearly suggests that processive antitermination is a common regulatory mechanism in GTAs, which evolved independently in two types of GTA within Alphaproteobacteria. To date, five types of GTAs have been identified in both bacteria and archaea. Given their phage heritage, it remains unclear whether other GTA types also depend on processive antitermination for regulation, and this should be explored in future studies.

Materials and Methods

Bacterial Strains and growth conditions

All bacterial strains, plasmids and oligonucleotides used in this study are listed in Supplementary Tables 1, 2 and 3, respectively. B. henselae strains were grown at 35°C on Columbia blood agar (CBA) supplemented with 5% sheep blood, appropriate antibiotics (gentamicin (Gm) – 10 μg/ml, kanamycin (Km) – 30 μg/ml, streptomycin (Sm) – 100 μg/ml, spectinomycin (Sp) – 50 μg/ml) and other supplements (isopropyl β-D-thiogalactosidase (IPTG) – 500 μM, anhydrotetracycline (ATc) – 10 ng/ml or 25 ng/ml) in a humidified atmosphere containing 5% CO2. B. henselae strains were stored as cryostocks at -80°C and prior the experiment were seeded on CBA for 3 days and subsequently expanded and grown on fresh plates for 2 days. E. coli strains were stored as cryostocks at -80°C and were grown at 37°C overnight in lysogeny broth (LB) or on solid LB agar plates (LA) supplemented with appropriate antibiotics (ampicillin (Ap) – 100 μg/ml, Gm – 10 μg/ml, Km – 30 μg/ml) and other supplements (2,6-diaminopimelic acid (DAP) – 1 mM, 1% glucose).

Construction of strains

All bacterial strains used in this study are listed in Table 1.

DNA manipulations were performed according to standard techniques and all cloned inserts were DNA sequenced to confirm sequence integrity. Chromosomal insertions or deletions in B. henselae were generated by a two-step gene replacement procedure as described in Schulein and Dehio (Schulein and Dehio, 2002). Plasmids were introduced into B. henselae strains by conjugation with E. coli strain JKE201 (MFDpir) using two-parental mating as described previously (Harms et al., 2017).

Construction of plasmids

All plasmids used in this study are listed in Table 2 and all oligonucleotides used in this study are listed in Table 3. Plasmids generated in this study were constructed as follows:

pQN011: Plasmid pTR1000 was digested with the restriction enzyme BamHI. Homologous fragments 1 (upstream) and 2 (downstream) were PCR amplified from heat-killed cells B. henselae with primers prQN029 and prQN030, prQN031 and prQN032, respectively. Those two fragments were cloned together by performing overlap extension PCR with primers prQN030 and prQN032. The fragment and the backbone were fused with the 5X In-Fusion® HD Enzyme Premix kit.

pQN018: brrG was PCR amplified from heat-killed cells B. henselae with primers prQN053 and prQN054. tetR was PCR amplified with primer prAK227 and prAK208 from 705-bp tetR-Ptet promoter template ordered as a synthetic fragment from Genescript. The PCR product generated by amplifying pBZ485_a_empty with prAK089 and prAK109 was used as a backbone. Three fragments were combined with the 5X In-Fusion® HD Enzyme Premix kit.

pAK126: brrG was PCR amplified from heat-killed cells B. henselae with primers prAK440 and prAK441. tetR was PCR amplified with primer prAK444 and prAK208 from 705-bp tetR-Ptet promoter template ordered as a synthetic fragment from Genescript. The PCR product generated by amplification of pPG100 with prAK116 and prAK167 was used as a backbone. Three fragments were combined with the 5X In-Fusion® HD Enzyme Premix kit.

pQN003: Fragment Pf3 was PCR amplified from heat-killed cells B. henselae with primers prQN005 and prQN014. dsRed was PCR amplified from pJC44 with primer prAK097 and prAK110. The PCR product generated by amplifying pBZ485_a_empty with prAK089 and prAK109 was used as a backbone. Three fragments were combined with the 5X In-Fusion® HD Enzyme Premix kit.

pQN006: gfpmut2 was PCR amplified from plasmid pCD353 with primers prQN015 and prQN016. The PCR product generated by amplifying pBZ485_a_empty with prAK089 and prAK109 was used as a backbone. Two fragments were combined with the 5X In-Fusion® HD Enzyme Premix kit.

pQN008: Plasmid pQN006 was digested with BamHI. Fragment sF1 was PCR amplified form heat-killed cells B. henselae with primers prQN021 and prQN022. The fragment and the backbone were ligated with the 5X In-Fusion® HD Enzyme Premix kit.

pQN009: Plasmid pQN006 was digested with BamHI. Fragment sF2 was PCR amplified form heat-killed cells B. henselae with primers prQN018 and prQN017. The fragment and the backbone were ligated with the 5X In-Fusion® HD Enzyme Premix kit.

pQN010: Plasmid pQN006 was digested with BamHI. Fragment sF3 was PCR amplified form heat-killed cells B. henselae with primers prQN019 and prQN020. The fragment and the backbone were ligated with the 5X In-Fusion® HD Enzyme Premix kit.

pQN020: The minimal predicted PGTA promoter was introduced into plasmid pQN006 by PCR with the primers prQN061 and prQN062. The plasmid was circularized with the 5X In-Fusion® HD Enzyme Premix kit.

pQN057: The PCR product generated by amplifying pQN006 with prQN111 and prAK109 was used as a backbone. F1 was PCR amplified from heat-killed cells B. henselae with primers prQ134 and prQN135. F1 and the backbone were combined with the 5X In-Fusion® HD Enzyme Premix kit.

pQN058: The PCR product generated by amplifying pQN006 with prQN111 and prAK109 was used as a backbone. F2 was PCR amplified from heat-killed cells B. henselae with primers prQ134 and prQN136. F2 and the backbone were ligated with the 5X In-Fusion® HD Enzyme Premix kit.

pQN059: The PCR product generated by amplifying pQN006 with prQN111 and prAK109 was used as a backbone. F3 was PCR amplified from heat-killed cells B. henselae with primers prQ134 and prQN137. F3 and the backbone were ligated with the 5X In-Fusion® HD Enzyme Premix kit.

pQN060: The PCR product generated by amplifying pQN006 with prQN111 and prAK109 was used as a backbone. F4 was PCR amplified from heat-killed cells B. henselae with primers prQ134 and prQN138. F4 and the backbone were ligated with the 5X In-Fusion® HD Enzyme Premix kit.

pQN061: The PCR product generated by amplifying pQN006 with prQN111 and prAK109 was used as a backbone. F5 was PCR amplified from heat-killed cells B. henselae with primers prQ134 and prQN139. F5 and the backbone were ligated with the 5X In-Fusion® HD Enzyme Premix kit.

pQN007: Fragment F6 was PCR amplified from heat-killed cells B. henselae with primers prQN005 and prQN014. Plasmid pQN006 and F6 were restricted with BamHI and Xmal and ligated with RAPID DNA Dephos and Ligation Kit.

pQN047: The mutated -10-like sequence was introduced into plasmid pQN007 by PCR with primers prQN120 and prQN121. The plasmid was circularized with the 5X In-Fusion® HD Enzyme Premix kit.

pQN048: N’-3x-FLAG was introduced into pQN018 by PCR with primers prQN114 and prQN115. The plasmid was circularized with the 5X In-Fusion® HD Enzyme Premix kit.

pQN049: C’-3x-FLAG was introduced into pQN018 by PCR with primers prQN116 and prQN117. The plasmid was circularized with the 5X In-Fusion® HD Enzyme Premix kit.

pAK180: Terminator TBBa_1006 was introduced into pQN058 by PCR with primers prAK592 and prAK593. The linear PCR product was used for the direct transformation of E. coli.

Measuring expression of the BaGTA locus with flow cytometry

B. henselae strains were grown as described above. Bacteria were resuspended in M199 supplemented with 10% FCS and other additives, e.g inducer, if required (hereafter M199+), diluted to OD600 nm = 0.008 and incubated in 24-well plate for a defined amount of time in a humidified atmosphere at 35°C and 5% CO2. After a defined amount of time bacteria were collected, centrifuged, and washed once with PBS. Bacteria were resuspended in 4% PFA and fixed for 7 min at RT. The fixation was stopped by adding PBS supplemented with 2% FCS. Samples were kept at 4°C in the dark before flow cytometric analysis. GFP fluorescence in samples was recorded with the BD LSRFortessa™ Cell Analyzer at the FACS Core Facility, Biozentrum, the University of Basel. Data were analysed using software FlowJo (Version 10.8.1). Visualisation of the data was done with GraphPad prism 8.

GTA transfer assay

GTA transfer assays were done as co-cultivations of two B. henselae strains. B. henselae strains were grown as described above. Bacteria were resuspended in M199+ supplemented, if required, with inducer at specified concentrations and diluted to OD600 nm = 0.008. For tested co-cultivation pair 1ml of each strain was mixed and a mixture was incubated in 24-well plate for 24h in a humidified atmosphere at 35°C and 5% CO2. After 24h bacteria were collected and centrifuged. Bacterial pellets were resuspended in 30ul of M199+ and resuspensions were used for serial dilutions followed by plating onto CBA plates supplemented with SmGm or SmGmSp antibiotics. Bacteria were grown for 6-15 d in a humidified atmosphere at 35°C and 5% CO2. Bacterial colonies were counted, and CFU/ml was determined for each co-cultivation pair. Frequency of double resistant bacteria (DRB) was calculated as the ratio of CFUs of recipient DRB (SmR GmR SpR) / total number of recipients (SmR GmR), yielding the transfer frequency per recipient cell.

Chromatin immunoprecipitation coupled with high throughput sequencing

B. henselae strains were grown as described above. Bacteria were harvested from plates and resuspended in 5 ml of Schneider’s Insect Medium (24.5 g/l Schneider’s Drosophila Powder Medium Revised, 5% (wt/vol) sucrose, 0.025 M HEPES, adjusted pH 7.25) supplemented with 10% FCS (hereafter Schneider’s Medium) (Guy et al., 2013; Riess et al., 2008). Bacteria were diluted to OD600 nm = 0.01 in 500 ml Schneider’s Medium for each strain and grown at 35°C with shaking 135 rpm for 4d until OD600 nm = 0.6-0.7. Bacteria were pelleted and pellets were resuspended in 50ml PBS. Bacteria were fixed with PFA with final concentration 1% and incubated for 10 min at RT. The fixation was stopped by adding glycine to a final concentration of 125 mM, and mixture was incubated for 20 min on ice. Fixed bacteria were centrifuged for 30 min at 4°C at 2000 x g, and bacterial pellets were frozen at -80°C. Thawed pellets were resuspended in 10 ml of ChIP Buffer (1.1% Triton X-100, 1.2 mM EDTA, 16.7 mM Tris-HCl (pH 8.1), 167 mM NaCl, supplemented with cOmplete™Mini EDTA-free Protease Inhibitor Cocktail) and cells were lysed by 4 passages with French Press. The lysate was sonicated, and DNA size was checked with the 1.5% agarose gel. The lysate was centrifuged with the maximal speed and supernatant was collected. A 50 ul aliquot of supernatant was taken as “Input DNA” for the subsequent sequencing and data analysis. Following manufacturers’ protocols, remaining supernatant was immunoprecipitated ON with anti-FLAG Magnetic Agarose or anti-HA magnetic beads for 3xFLAG-tagged and HA-tagged proteins, respectively. Magnetic beads/agarose were immobilized and after removal of supernatant were sequentially rinsed with a panel of buffers: Low Salt Wash Buffer (0.1% SDS, 1% Triton X-100, 2 mM EDTA, 20 mM Tris-HCl (pH 8.1), 150 mM NaCl), High Salt Wash Buffer (0.1% SDS, 1% Triton X-100, 2 mM EDTA, 20 mM Tris-HCl (pH 8.1), 500 mM NaCl), LiCl Wash Buffer (250 mM LiCl, 1% NP-40, 1% deoxycholate, 1 mM EDTA, 10 mM Tris-HCl (pH 8.1)), TBS Buffer (50 mM Tris-HCl, 150 mM NaCl). The thawed Input control and immobilized Magnetic beads/agarose were mixed with 450 ul of TBS buffer, 15 µL of 5 M NaCl and 0.5 ul RNAse A (100mg/ml). The mixtures were incubated at 65°C ON. Then 4 µL of 0.5 M EDTA and 2 µL of 10 mg/mL Proteinase-K were added and the mixtures were incubated at 45°C for 2 h. The resulting mixtures were thoroughly mixed in 1:1 ratio with phenol/chlorophorm/isoamylalcohol combination and centrifuged at >10 000g for 1 min. The upper aqueous phase was transferred to a new tube, mixed in 1:1 ratio with chloroform and centrifuged at >10’000 x g for 1 min. The upper aqueous phase was transferred to a new tube and mixed with: (i) 25 ug of linear polyacrylamide; (ii) NaAc to a final concentration of 0.3 M; (iii) in 1:1 ratio with 100% isopropanol. The solution was centrifuged at >10’000 x g for 30min, washed twice with cold 70% EtOH. Pelleted DNA was dried and resuspend in water ON. Library preparation and sequencing were done at the Genomics Facility Basel, D-BSSE, ETH-Zurich.

Bioinformatic analysis

All alignments were done with the Clustal Omega and MUSCLE implemented in Geneious Prime 2020.2.3 (Edgar, 2004; Sievers et al., 2011). Mapping of ChIP-Seq data was done with bowtie2 (Langmead and Salzberg, 2012). Converting SAM to BAM, extracting unique binding reads and sorting were done with SAMtools (Danecek et al., 2021). Coverage plots were generated with bamCoverage and visualized with the Integrative Genomics Viewer (Ramírez et al., 2016; Robinson et al., 2011). Promoter predictions were done with BProm (Solovyev et al., 2011). Models of protein structures were either done with AlphaFold2 or taken from AlphaFold Protein Structure Database (Jumper et al., 2021; Varadi et al., 2022). Comparing the predicted models to Protein Data Bank (PDB) was done with the Dali server (Holm, 2022). Predictions of the secondary structure of the putative terminator was done with Mfold with the default parameters (Zuker, 2003). Calculations were performed at sciCORE (http://scicore.unibas.ch/) scientific computing center at the University of Basel.

eLife Assessment

This manuscript describes an important finding of the transcriptional control of a chimeric gene transfer agents (GTA) cluster in Bartonella by a processive anti-termination factor (BrrG). The evidence provided is solid. This manuscript will interest researchers working on transcriptional regulation, horizontal gene transfer, and phages.

Reviewer #1 (Public review):

Summary:

Gene transfer agent (GTA) from Bartonella is a fascinating chimeric GTA that evolved from the domestication of two phages. Not much is known about how the expression of the BaGTA is regulated. In this manuscript, Korotaev et al noted the structural similarity between BrrG (a protein encoded by the ror locus of BaGTA) to a well-known transcriptional anti-termination factor, 21Q, from phage P21. This sparked the investigation into the possibility that BaGTA cluster is also regulated by anti-termination. Using a suite of cell biology, genetics, and genome-wide techniques (ChIP-seq), Korotaev et al convincingly showed that this is most likely the case. The findings offer the first insight into the regulation of GTA cluster (and GTA-mediated gene transfer) particularly in this pathogen Bartonella. Note that anti-termination is a well-known/studied mechanism of transcriptional control. Anti-termination is a very common mechanism for gene expression control of prophages, phages, bacterial gene clusters, and other GTAs, so in this sense, the impact of the findings in this study here is limited to Bartonella.

Strengths:

Convincing results that overall support the main claim of the manuscript.

Weaknesses:

A few important controls are missing.

Reviewer #2 (Public review):

Summary:

In this study, the authors identified and characterized a regulatory mechanism based on transcriptional anti-termination that connects the two gene clusters, capsid and run-off replication (ROR) locus, of the bipartite Bartonella gene transfer agent (GTA). Among genes essential for GTA functionality identified in a previous transposon sequencing project, they found a potential antiterminatior of phage origin within the ROR locus. They employed fluorescence reporter and gene transfer assays of overexpression and knockout strains in combination with ChiPSeq and promoter-fusions to convincingly show that this protein indeed acts as an antiterminator counteracting attenuation of the capsid gene cluster expression.

Impact on the field:

The results provide valuable insights into the evolution of the chimeric BaGTA, a unique example of phage co-domestication by bacteria. A similar system found in the other broadly studied Rhodobacterales/Caulobacterales GTA family suggests that antitermination could be a general mechanism for GTA control.

Strengths:

Results of the selected and carefully designed experiments support the main conclusions.

Weaknesses:

It remains open why overexpression of the antiterminator does not increase the gene transfer frequency.

Author response:

The following is the authors’ response to the original reviews

Reviewer 1 (Public review):

Summary:

Gene transfer agent (GTA) from Bartonella is a fascinating chimeric GTA that evolved from the domestication of two phages. Not much is known about how the expression of the BaGTA is regulated. In this manuscript, Korotaev et al noted the structural similarity between BrrG (a protein encoded by the ror locus of BaGTA) to a well-known transcriptional anti-termination factor, 21Q, from phage P21. This sparked the investigation into the possibility that BaGTA cluster is also regulated by anti-termination. Using a suite of cell biology, genetics, and genome-wide techniques (ChIP-seq), Korotaev et al convincingly showed that this is most likely the case. The findings offer the first insight into the regulation of GTA cluster (and GTA-mediated gene transfer) particularly in this pathogen Bartonella. Note that anti-termination is a well-known/studied mechanism of transcriptional control. Anti-termination is a very common mechanism for gene expression control of prophages, phages, bacterial gene clusters, and other GTAs, so in this sense, the impact of the findings in this study here is limited to Bartonella.

Strengths:

Convincing results that overall support the main claim of the manuscript.

Weaknesses:

A few important controls are missing.

We sincerely appreciate reviewer #1's positive assessment of our manuscript. In response to the concern regarding control samples/experiments, we have addressed this issue in our revision, by providing data of the replicates of our experiments. We acknowledge that antitermination is a well-established mechanism of expression control in bacteria, including bacterial gene clusters, phages, prophages, and at least one other GTA. As reviewer #2 also noted, our study presents a unique example of phage co-domestication, where antitermination integrates both phage remnants at the regulatory level. We have emphasized this original aspect more clearly in the revised manuscript.

Reviewer 1 (Recommendations for the authors):

(1) Provide Rsmd and DALI scores to show how similar the AlphaFold-predicted structures of BrrG are to other anti-termination factors. This should be done for Fig1B and also for Suppl. Fig 1 to support the claim that BrrG, GafA, GafZ, Q21 share structural features.

In the revised manuscript we provide Rsmd and DALI scores in the supplementary Fig. 1A (Suppl. Fig. 1A). In Suppl. Fig. 1B we further include a heatmap of similiarity values.

(2) Throughout the manuscript, flow cytometry data of gfp expression was used and shown as single replicate. Korotaev et al wrote in the legends that error bars are shown (that is not true for e.g. Figs. 3, 4, and 5). It is difficult for reviewers/readers to gauge how reliable are their experiments.

In the revised manuscript we show all replicates for the flow cytometry histograms.

For Fig. 2C, all replicates are provided in Suppl. Fig. 3.

For Fig. 3B, all replicates are provided in Suppl. Fig. 4.

For Fig. 4B, all replicates are provided in Suppl. Fig. 5.

For Fig. 5B, all replicates are provided in Suppl. Fig. 6.

(3) I am unsure how ChIP-seq in Fig. 2A was performed (with anti-FLAG or anti-HA antibodies? I cannot tell from the Materials & Methods). More importantly, I did not see the control for this ChIP-seq experiment. If a FLAG-tagged BrrG was used for ChIP-seq, then a WT non-tagged version should be used as a negative control (not sequencing INPUT DNA), this is especially important for anti-terminator that can co-travel with RNA polymerase. Please also report the number of replicates for ChIP-seq experiments.

Fig. 2A presents the coverage plot from the ChIP-Seq of ∆brrG +pPtet:3xFLAG-brrG (N’ in green). As anticipated by the referee, we had used ∆brrG +pTet:brrG (untagged) as control (grey). Each strain was tested in a single replicate. The C-terminal tag produced results similar to the untagged version, suggesting it is non-functional. All tested tags are shown in Supplementary Figure 2.

(4) Korotaev et al mentioned that BrrG binds to DNA (as well as to RNA polymerase). With the availability of existing ChIP-seq data, the authors should be able to locate the DNA-binding element of BrrG, this additional information will be useful to the community.

We identified a putative binding site of BrrG using our ChIP-Seq data. The putative binding site is indicated in Fig. 2D of the revised manuscript.

(5) Mutational experiments to break the potential hairpin structure are required to strengthen the claim that this putative hairpin is the potential transcriptional terminator.

We did not claim the identified hairpin is a confirmed terminator, but proposed it as a candidate. We agree with the referee that the suggested experiment would be necessary to definitively establish its function. However, our main objective was to show that BrrG acts as a processive terminator, which we demonstrated by replacing the putative terminator with a well-characterized synthetic one that BrrG successfully bypassed. Therefore, we chose not to perform the proposed experiment and have accordingly softened our conclusions regarding the hairpin’s potential terminator function.

Reviewer 2 (Public review):

Summary:

In this study, the authors identified and characterized a regulatory mechanism based on transcriptional anti-termination that connects the two gene clusters, capsid and run-off replication (ROR) locus, of the bipartite Bartonella gene transfer agent (GTA). Among genes essential for GTA functionality identified in a previous transposon sequencing project, they found a potential antiterminatior of phage origin within the ROR locus. They employed fluorescence reporter and gene transfer assays of overexpression and knockout strains in combination with ChiPSeq and promoter-fusions to convincingly show that this protein indeed acts as an antiterminator counteracting attenuation of the capsid gene cluster expression.

Impact on the field:

The results provide valuable insights into the evolution of the chimeric BaGTA, a unique example of phage co-domestication by bacteria. A similar system found in the other broadly studied Rhodobacterales/Caulobacterales GTA family suggests that antitermination could be a general mechanism for GTA control.

Strengths:

Results of the selected and carefully designed experiments support the main conclusions.

Weaknesses:

It remains open why overexpression of the antiterminator does not increase the gene transfer frequency.

We are grateful for reviewer #2's thoughtful and encouraging feedback on our manuscript. The reviewer raises an important question about why overexpression of the antiterminator does not increase gene transfer frequency. While we acknowledge this point, we consider it beyond the scope of the current study. Our findings clearly demonstrate that the antiterminator induces capsid component expression in a large proportion of cells. However, the fact that this expression plateaus at high levels rather than exhibiting a transient peak, as seen in the wild type, suggests that antiterminators do not regulate GTA particle release via lysis. We are actively investigating this further through additional experiments, which we plan to publish separately from this study.

Reviewer 2 (Recommendations for the authors):

(1) The authors wrote "GTAs are not self-transmitting because the DNA packaging capacity of a GTA particle is too small to package the entire gene cluster encoding it" (page 3). I thought that at least the Bartonella capsid gene cluster should be self-transmissible within the 14 kb packaged DNA (https://doi.org/10.1371/journal.pgen.1003393, https://doi.org/10.1371/journal.pgen.1000546). This was also concluded by Lang et al (https://doi.org/10.1146/annurev-virology-101416-041624). In this case the presented results would have important implications. As the gene cluster and the anti-terminator required for its expression are separated on the chromosome, it would not be possible to transfer an active GTA gene cluster, although the DNA coding for the genes required for making the packaging agent itself, theoretically fits into a BaGTA particle. Could the authors comment on that? I think it would be helpful to add the sizes of the different gene clusters and the distance between them in Fig. 2A. The ROR amplified region spans 500kb, is the capsid gene cluster within this region?

We thank the reviewer for bringing up this interesting point. The ror gene cluster, which encodes the antiterminator BrrG, is approximately 9.2 kb in size and could feasibly be packaged in its entirety into a GTA particle. In contrast, the bgt cluster (capsid cluster) is approximately 20 kb in size —exceeding the packaging limit of GTA particles—and is separated from the bgt cluster by approximately 35 kb. Consequently, if the ror cluster is transferred via a GTA particle into a recipient host that does not encode the bgt gene cluster, the ror cluster would not be expressed.

We added the sizes of the gene clusters to Fig. 1A.

(2) Another side-note regarding the introduction: On page three the authors write: "GTAs encode bacteriophage-like particles and in contrast to phages transfer random pieces of host bacterial DNA". While packaging is not specific, certain biases in the packaging frequency are observed in both studied GTA families. For Bartonella this is ROR. In the two GTA-producing strains D. shibae and C. crescentus origin and terminus of replication are not packaged and certain regions are overrepresented (https://doi.org/10.1093/gbe/evy005, https://doi.org/10.1371/journal.pbio.3001790). Furthermore, D. shibae plasmids are not packaged but chromids are. I think the term "random" does not properly describe these observations. I would suggest using "not specific" instead.

We thank the reviewer for this suggestion and adjusted the wording on p. 3 accordingly.

(3) Page 5: Remove "To address this". It is not needed as you already state "To test this hypothesis" in the previous sentence.

We adjusted the working on p.5 accordingly.

(4) I think the manuscript would greatly benefit from a summary figure to visualize the Q-like antiterminator-dependent regulatory circuit for GTA control and its four components described on pages 15 and 16.

We thank the reviewer for this valuable suggestion. We included a summary figure (Fig. 6) in the discussion section of the revised manuscript.

(5) Page 17: It might be worth noting that GafA is highly conserved along GTAs in Rhodobacterales (https://doi.org/10.3389/fmicb.2021.662907) and so is probably regulatory integration into the ctrA network (https://doi.org/10.3389/fmicb.2019.00803). It's an old mechanism. It would be also interesting to know if it is a common feature of the two archetypical GTAs that the regulator is not part of the cluster itself.

We agree with the reviewer’s comments and have revised the wording to state that GafA is highly conserved.

Version 2

Abstract

Abstract

Gene Transfer Agents (GTAs) are mobile genetic elements derived from bacteriophages that mediate genome-wide horizontal gene transfer (HGT) in diverse groups of prokaryotes. BaGTA, encoded by all the pathogens of the genus Bartonella, is a chimeric GTA that evolved by the domestication of two phages. The run-off-replication module ROR of one phage is integrated with the capsid production, DNA packaging and lysis machinery Bgt of a second phage. Restricted to a self-sacrificing subset of the bacterial population, the position-specific DNA amplification and packaging of a genomic plasticity region enriched for genes involved in host interaction and adaptation selectively enhances the HGT frequency of these pathogenicity genes. This feature of BaGTA is considered a key innovation underlying the evolutionary success of Bartonella. Little is known, however, about the mechanism mediating the coordinated expression of the ror and bgt loci. Here, we established the regulatory hierarchy, with ror acting upstream of the capsid gene cluster bgtA-K. BrrG, encoded by the ror locus, controls the transcription of the bgtA-K operon by functioning as a processive antiterminator. This study provides the first insights into the mechanism controlling the coordinated expression of the two BaGTA modules of divergent phage origin. Beyond BaGTA, we propose that antitermination is a broadly relevant mechanism for controlling HGT by GTAs of the Alphaproteobacteria.

Introduction

Timed and tightly controlled gene expression is crucial in bacterial cells. A variety of mechanisms to control transcription are known, comprising both positive and negative regulators (reviewed in Turnbough, 2019). One of the mechanisms of negative regulation involves transcription attenuation at termination sites (reviewed in Santangelo and Artsimovitch, 2011). The regulation of transcription attenuation is achieved by a diverse class of antiterminators which counteract the attenuation (Santangelo and Artsimovitch, 2011). Antiterminators fall into two categories and can either act specifically at a single, defined terminator site or convert RNA polymerase into a termination-resistant complex that is generally resistant to multiple termination sites over a longer distance (Santangelo and Artsimovitch, 2011; reviewed in Goodson and Winkler, 2018). In the latter case, such antiterminators are referred to as processive antiterminators. Well-known examples of processive antiterminators are Q and N of bacteriophage lambda which control progression of the infection cycle (Goodson and Winkler, 2018).

Peculiar examples of phage evolution are Gene Transfer Agents (GTAs) (reviewed in Québatte and Dehio, 2019; Banks and Tung, 2024; Craske et al., 2024). GTAs are mobile genetic elements, which emerged from defective temperate bacteriophages. GTAs encode bacteriophage-like particles and in contrast to phages transfer unspecific pieces of host bacterial DNA. Despite their phage origin, GTAs are not self-transmitting because the DNA packaging capacity of a GTA particle is too small to package the entire gene cluster encoding it. Therefore, GTAs are transferred vertically (Banks and Tung, 2024; Craske et al., 2024; Kogay et al., 2022; Lang et al., 2012). GTAs are found both in bacteria and archaea, and are one of the non-canonical mechanisms of horizontal gene transfer (HGT) (Arnold et al., 2022).

Bacterial pathogens of the genus Bartonella encode a chimeric GTA known as BaGTA (Québatte and Dehio, 2019; Berglund et al., 2009; Guy et al., 2013). BaGTA encodes the gene clusters bgtA-K, responsible for the synthesis and assembly of GTA particles, bgtL-R, with orphan functions, and bgtS-V, involved in putative host cell lysis (Fig. 1A). DNA packaging in GTA particles is coupled to a process of site-specific DNA amplification mediated by the ror locus located in between the bgtL-R and bgtS-V gene clusters. This locus, emerging from a different phage domestication event, includes an origin of run-off replication (ROR) and the associated brr gene cluster (brrA-G) (Berglund et al., 2009; Guy et al., 2013; Québatte et al., 2017; Tamarit et al., 2018) (Fig. 1A). The bgtA-K and ror loci are located within a high plasticity zone of the genome that is prone to DNA integrations and is highly enriched for genes encoding virulence properties involved in host interaction and adaptation (Berglund et al., 2009; Guy et al., 2013). This specific genome architecture leads to ROR-driven amplification and packaging of these pathogenicity genes, resulting in elevated transfer rates to other Bartonella bacteria compared to housekeeping genes encoded more distantly to the ROR origin (Québatte et al., 2017). This process is considered to provide the recombination substrates for the duplication and evolutionary diversification of genes encoding the type IV secretion systems (T4SSs) and cognate effector proteins which collectively represent a key innovation associated with the adaptive radiation of these pathogens to diverse mammalian hosts (Guy et al., 2013; Engel et al., 2011). Additionally, it is also considered to facilitate the fast evolution of surface-located autotransporters serving in host interaction and adaptation, such as the recombination-driven antigenic variation of the CFA autotransporter (Siewert et al., 2022). Thus, BaGTA is a crucial factor for the fitness and adaptability of pathogens in the Bartonella genus (Québatte and Dehio, 2019). However, little is known about how the expression of the bgt and brr loci derived from separate phages is regulated in a coordinated manner.

BrrG encodes a putative antiterminator essential for bgtA-K expression and BaGTA-mediated gene transfer.

(A) Schematic representation of the BaGTA locus. Upper part: Structure and size of the bgtA-K and ror gene clusters. Lower part: Close-up of the bgtA-K locus encoding capsid and packaging functions and of the ror locus, encoding the run-off-replication origin oriROR and the associated brrA-G gene cluster. For the reporter strain bgtC-gfp and its derivatives used in D-G, the insertion site of the transcriptional gfp reporter within the bgtA-K locus is indicated. (B) Comparison of the structure of Q21 antiterminator (6p1b) of phage P21 and Alphafold2 model of BrrG (A0A0R4J952). Similar structural features are presented in identical colour. (C) Schematic presentation of (i) the flow cytometry assay to monitor expression of the bgtA-K locus using the bgtC-gfp transcriptional reporter strain (upper panel) and (ii) the assay to monitor GTA transfer via transfer of GmR from the GTA-positive donor strain to the ΔbgtCD SpR GTA-negative (ΔGTA) recipient strain resulting in SpRGmRbacteria and formula to calculate the frequency of double resistant bacteria (fDRB) as a proxy for GTA transfer. (D) BrrG overproduction increases the bacterial population expressing bgtA-K. BrrG expression was induced in strain bgtC-gfp (pPtet-brrG) with anhydrotetracycline (aTc). Strain bgtC-gfp H1GmR (bgtC-gfp BH01490::Himar1GmR) displaying wild-type properties served as a positive control. (E) BrrG overproduction does not enhance GTA transfer. Strain bgtC-gfp H1GmRdisplaying wild-type properties and the GTA-negative mutant ΔbgtCD H1GmR(ΔbgtCD BHRS00695::Himar1GmR) served as a positive and negative control, respectively. (F) Deletion of brrG (ΔbrrG bgtC-gfp H1GmR [ΔbrrG BH01490::Himar1GmR]) results in reduced expression of the bgtA-K locus. (G) Overproduction of BrrG in a ΔbrrG mutant background restores GTA transfer frequency to the wild-type level. BrrG expression was induced in strain ΔbrrG H1GmR(pPtet-brrG) with aTc. (D-G) The time-course experiments were started with the incubation of CBA agar-grown bacteria in M199 medium supplemented with 10% FCS. 25 ng/ml aTc was added where indicated to induce expression by the Ptet promoter. The experiments were performed in biological triplicates (D, F, G) or quadruplicates (E), error bars indicate SD.

In this study, we elucidated the regulatory hierarchy between the bgtA-K and ror loci, resolving the mechanism of gene expression control. We discovered that BaGTA operates under a strict regulatory framework, with a protein encoded by the ror locus and controlling the expression of the bgtA-K capsid-production locus. Specifically, our research revealed that in absence of BaGTA activation, transcription of the bgtA-K locus is attenuated by a putative terminator site. Upon BaGTA activation by a yet unknown process, the ror locus-encoded BrrG overcomes this transcription attenuation of the bgtA-K locus by acting as a processive antiterminator. Furthermore, we propose that antitermination is a common mechanism of GTA regulation in the Alphaproteobacteria. Collectively, these findings offer significant insights into the regulation of BaGTA-mediated gene transfer in pathogens of the genus Bartonella and suggest a broader impact of processive antitermination as a regulatory principle in related GTAs.

Results

BrrG shows structural similarity to the P21 phage-encoded Q21 antiterminator and is essential for GTA transfer and expression of capsid components

Previously, we identified by multiplexed transposon sequencing the genes of BaGTA encoding both for donor transfer and recipient uptake functions (Québatte et al., 2017). However, many of the identified genes had only hypothetical or no functional sequence-based annotation, leaving their contribution to BaGTA function unclear. To address this knowledge gap, we employed the protein structure prediction tool AlphaFold2 to model unannotated proteins of BaGTA genes and searched the Protein Data Bank (PDB) database for similar structures. We discovered that the brrG gene (Fig. 1A), which is located in the ror locus, encodes a protein with structural similarity to the antiterminator Q21 of the lambdoid bacteriophage P21 (Fig. 1B). Q21 and its counterpart Q from phage λ (Qλ) function as processive antiterminators involved in bypassing intrinsic terminators (Guo et al., 1991; Grayhack and Roberts, 1982). This enables uninterrupted expression of the late genes, which encode capsid morphogenesis and lysis cassettes. Therefore, we hypothesized that, similar to Q21 and Qλ, BrrG represents a putative antiterminator controlling gene expression within the BaGTA regulatory network.

To test this hypothesis, we assessed the effect of brrG overexpression on the transcription of the bgtA-K locus that encodes components of the capsid synthesis and assembly machinery as essential components of BaGTA production. We utilized the previously established reporter strain bgtC-gfp that contains a promoter-less gfpmut2 inserted in the bgtA-K locus in the intergenic region between bgtC and bgtB (Québatte et al., 2017) (Fig. 1A). This transcriptional reporter enabled us to monitor bgtA-K expression over time using flow cytometry (Fig. 1C, upper panel). We observed that plasmid-based overproduction of BrrG increased the fraction of GFP fluorescent bacteria up to 75%, compared to about 15% at native BrrG expression (Fig. 1D), thus massively expanding the fraction of cells expressing the bgtA-K locus. To test whether this increase is followed by GTA production and elevated DNA transfer, we performed the GTA transfer assay. For that, we cocultivated donor and recipient cells carrying different resistance markers followed by plating on selective media to enumerate double-resistant bacteria as a read-out for the frequency of GTA-mediated gene transfer (Fig. 1C, lower panel). We observed that - while the proportion of cells expressing the BaGTA locus increased as a result of BrrG overproduction, GTA transfer rates remained similar to native BrrG expression (Fig. 1E). In contrast, deleting brrG led to a significant decrease in the proportion of fluorescent bacteria (Fig. 1F) and caused DNA transfer rates to drop at least 100-fold compared to those observed at native BrrG expression (Fig. 1G). Together, these findings indicate that BrrG plays a crucial role in regulating BaGTA. It is capable of activating the expression of capsid components in a large part of the population that is initially GTA inactive. However, overproduction of BrrG is not sufficient to increase the rate of DNA transfer, indicating that it acts in concert with other regulators controling other crucial steps of this complex process.

BrrG binds in proximity to the PbgtA-K promoter located upstream of to the putative bgtA-K operon

In lambdoid phages, both Q21 and Qλ are DNA-binding proteins that recognize a motif located in the promoter region of the late operon, which encodes lysis cassettes and proteins required for capsid morphogenesis (Guo et al., 1991; Grayhack and Roberts, 1982; Yin et al., 2019; Shi et al., 2019; Yin et al., 2022). Considering the structural similarity between BrrG and Q21, we hypothesized that BrrG may be functionally conserved and thus able to directly bind to DNA at the yet unidentified promoter of the bgtA-K locus. To test this, we performed chromatin immunoprecipitation of epitope-tagged BrrG followed by DNA sequencing. We found that BrrG binds to DNA, with a single binding site located in the intergenic region far upstream of the bgtK gene representing the first conserved gene of the bgtA-K locus (Fig. 2A, Suppl. Figure 2) (Guy et al., 2013; Tamarit et al., 2018). Given this remote upstream location of the BrrG binding site, we wondered whether the promoter elements might also be located nearby. First, we tested whether the 999 bp fragment spanning the region upstream of bgtK until the end of clpB could drive expression of a fused promoter-less dsRed gene using flow cytometry as readout (Fig. 2A, plasmid pPf3-dsRed). Strikingly, this fragment did not only demonstrate promoter activity but also displayed at the population level a temporal control resembling closely the expression pattern of the chromosomal transcriptional reporter strain bgtC-gfp H1GmR with a characteristic peak in the size of expressing population at the 6 h time-point of the 24 h expression kinetic (Fig. 2B). This finding suggests that the bgtA-K locus is organized as a transcriptional unit (operon) that is driven by a promoter localizing to the 999 bp fragment upstream of bgtK, which we thus named PbgtA-K (Fig. 2A).

BrrG binds in proximity to the PbgtA-K promoter located upstream of the putative bgtA-K operon.

(A) Location of the PbgtA-K promoter and of the BrrG binding site identified by ChIP-Seq (gray histogram). ΔbrrG + pPtet-brrG (no tag) is shown in dark grey; ΔbrrG + pPtet-3x-FLAG-brrG is shown in green. The fragments upstream of the bgtA-K locus were tested for promoter activity in promoter-probe plasmids as indicated. (B) The fusion of the 999 bp fragment spanning from the region between bgtK and clpB to dsRed (strain WT (PPf3-dsRed) displays a temporal transcriptional profile similar to the chromosomal transcription reporter for the bgtA-K locus (strain bgtC-gfp H1GmR [bgtC-gfp BH01490::Himar1GmR]). (C) The PbgtA-K promoter is located far upstream of the bgtA-K locus. Left: Temporal transcriptional profiles controlled by fragments spanning nucleotide 1-416 (plasmid psF1), nucleotides 271-782 (plasmid psF2), or nucleotides 594-999 (plasmid psF3) of the 999 bp fragment in Pf3. Right: Corresponding histograms of fluorescence distribution at the 6 h time-point. Strain bgtC-gfp displaying wild-type expression pattern of the bgtA-K locus was used as control. (D) DNA sequence alignment of the predicted promoter region and putative BrrG binding site within psF3 fragment with homologous sequences from various Bartonella species. The predicted -35 and -10 promoter elements and the inferred transcriptional start site (+1) are indicated. Numbers below indicate the location of the promoter elements relative to bgtK. (B, C) The time-course experiments were started with the incubation of CBA agar-grown bacteria in M199 medium supplemented with 10% FCS. The experiments were done in biological triplicates, error bars indicate SD.

Next, to narrow down the location of PbgtA-K, we split the 999 bp fragment into three partially overlapping fragments and tested them with the same approach using a gfpmut2-based promoter-probe plasmid (plasmids psF1, psF2 and psF3, Fig. 2A). We observed that the most upstream fragment spanning nucleotides 594-999 was able to drive broad expression of gfpmut2 at population level (plasmid psF3 in Figure 2C). The fraction of fluorescent bacteria was substantially higher compared to the chromosomal transcriptional reporter strain bgtC-gfp, expanding beyond 40% at each time-point (Fig. 2C, Suppl. Figure 3). Importantly, the lack of temporal regulation as seen by the chromosomal transcriptional reporter strain bgtC-gfp and likewise the 999 bp fragment in plasmid Pf3 (compare Figs. 2B and 2C) suggests that absence of downstream control elements located in the more proximal region spanning nucleotides 1-593 converted PbgtA-K into a constitutive state. To map PbgtA-K based on conserved elements (-10 and -35 boxes) we aligned corresponding sequences from a selection of Bartonella species covering the genomic diversity of this genus. By this comparative approach, we identified a highly conserved sequence motive (TTGACA-N17-TATCAT) that shows almost full identity to the canonical sigma-70 promoter sequence of E. coli (-35 box: 100%, -10 box: >80%, Fig. 2D) (Campbell et al., 2002; Murakami et al., 2002). This suggests that PbgtA-Kfunctions as a minimal promoter of the bgtA-K locus. Interestingly, the BrrG binding region coincides with PbgtA-K, suggesting that BrrG controls transcription either by direct binding to PbgtA-K or in the vicinity of it. Taken together, we identified PbgtA-K as a strong constitutive promoter located far upstream of the bgtA-K locus, driving its expression as a transcriptional unit along with more proximal elements required for controlling temporal expression at the population level. Additionally, we found that the putative antiterminator BrrG, similar to Q21 and Qλ, is a DNA-binding protein that directly binds to PbgtA-K and likely controls bgtA-K operon expression at population level.

A long leader sequence attenuating expression is located downstream of the PbgtA-K promoter

To confirm that the identified conserved sequence is the promoter and to hint at the function of the long downstream sequence in transcription, we generated a panel of promoter-probe plasmids harbouring truncations of the 999 bp fragment bearing promoter activity (pF0-gfp to pF6-gfp) (Fig. 3A). We observed that the 29 bp fragment spanning the predicted PbgtA-K displayed fluorescence in the vast majority of cells, confirming the promoter prediction (F0 fragment, Fig. 3B). Similar promoter activities were recorded for the fragments F1 and F2 that are extended by 27 bp or 133 bp towards the bgtA-K locus, respectively. A striking difference in fluorescence distribution, with an almost complete loss of fluorescent bacteria, was observed for the F3 fragment, which contains an additional 139 bp relative to F2 (Fig. 3B, Suppl. Figure 4). Fragments F4-F6 with further downstream extensions also displayed pattern of low fluorescence, while a bi-model distribution with low- and intermediate-expressing bacteria was observed for fragments F4 and F6.

A putative terminator is encoded in the long non-protein coding sequence downstream of the PbgtA-K promoter.

(A) Schematic representation of the 5’-part of the bgtA-K locus with the truncations tested with promoter-probe plasmids. Locations of the PbgtA-Kpromoter and the predicted terminator are shown. (B) Histograms of fluorescence distribution measured for the bacterial populations containing the promoter-probe plasmids with various truncations at 6 h. (C) Predicted secondary structure and free energy of the putative TbgtA-K terminator. Numbers in parentheses indicate the location of the terminator relative to bgtK. The time-course experiment was started with the incubation of CBA plate-grown bacteria in M199 medium supplemented with 10% FCS. Experiments were performed in biological triplicates.

Because fragment F3 is the shortest fragment displaying reduced expression, we hypothesized that a terminator could be present in the sequence extending the highly fluorescent F2 fragment. Indeed, a putative L-shaped terminator with a free energy (ΔG) of -6 kcal/mol is predicted within the stretch of 139 bp of the F3 fragment, which might explain the observed difference in expression levels (Fig. 3C). Altogether, these results demonstrate that the minimal promoter enables uniform and constitutive expression of the BaGTA locus at population level. The observed temporal regulation and heterogeneity of the gene cluster expression appear to be provided by the stretch of 864 bp located downstream of the promoter and upstream to bgtK as first gene of the bgtA-G locus. This stretch of DNA contains the putative terminator and could harbour other terminators or additional regulatory elements.

A -10-like element is required for expression of the bgtA-K locus

In phage lambda, the process of Q-mediated antitermination relies on the engagement of Q into the complex with RNA-polymerase (RNAP) and the conversion of the latter into a termination-resistant form (Perdue and Roberts, 2011). This process begins with RNAP pausing at the -10- like element, a six-nucleotide motif (NANNNT), located just downstream of the -10 promoter element, which allows Q to engage with the transcription complex (Perdue and Roberts, 2011).

Therefore, we set out to determine whether a -10-like element is present in the vicinity of the identified PbgtA-K minimal promoter. We examined the previously generated alignment and identified a conserved NANNNT site located 5 bp downstream of the identified promoter PbgtA-K(Fig. 4A). As the conserved nucleotides A and T at positions +2 and +6 in phage lambda and other phages are also conserved in this motif directly downstream of PbgtA-K, we hypothesized that this site could act as a -10-like element (Fig. 4A). To validate this hypothesis, we used the promoter-probe plasmid containing the F6 fragment (Fig. 3A), and introduced two point mutations, replacing the conserved A in position +2 to G and T in position +6 to C (pF6(-10mut)-gfp), respectively. Additionally, we introduced a second plasmid directing aTc-inducible BrrG overproduction to test whether these mutations would affect the activity of the putative antiterminator in the flow cytometry assay (Fig. 1C). In contrast to plasmid pF6-gfp encoding the native -10-like element (Fig. 4B, left, -aTc), pF6(-10mut)-gfp did not display the characteristic kinetics of GFP-expression of wild-type bacteria with a peak at 6 h (Fig. 4B, right, -aTc, Suppl. Figure 5). Moreover, neither did it display the massive expansion of the GFP-expressing population upon BrrG overproduction (+aTc) (Fig. 4B, Suppl. Figure 5). Thus, the replacement of two conserved nucleotides at the putative -10-like element completely abolished transcription of the downstream bgtA-K locus under both native BrrG expression and BrrG overproduction conditions. This suggests that the identified conserved site is a functional -10-like element essential for the BrrG-mediated antitermination.

The -10-like element downstream of the PbgtA-K promoter is crucial for the antitermination by BrrG.

(A) Alignment of Bartonella spp. consensus and lambdoid phage sequences (P21, phi80 and lambda) sequences containing PbgtA-K and the PR’ promoters followed by the -10-like elements, respectively. Promoter -35 and -10 elements, as well as the -10-like elements are marked. (B) Histograms of fluorescence distribution for the bacterial populations containing the promoter-probe plasmid pF6 (compare to Fig. 3A) with the PbgtA-Kpromoter and the adjacent native -10-like element or its derivative, plasmid pF6(-10mut), carrying mutations in two conserved positions of the -10-like element. Plasmid pPtet-brrG encodes an aTc-inducible brrG expression cassette. BrrG overproduction was induced by 25 ng/ml aTc. The time-course experiment was started with the incubation of CBA plate-grown bacteria in M199 medium supplemented with 10% FCS. Inducer aTc was added at time-point 0 h. Experiments were performed in biological triplicates.

BrrG is a processive antiterminator allowing RNAP to bypass the termination site

We showed that BrrG shares structural features with the processive antiterminator Q21 and that its function is dependent on the -10-like element suggesting that BrrG belongs to the class of processive antiterminators. To validate this hypothesis, we tested the antitermination properties of BrrG towards a synthetic terminator. For this, we used the artificially designed strong terminator BBa_B1006 from the iGEM collection and placed it in between the F2 fragment (see Fig. 3A), which contains the native PbgtA-K promoter and downstream -10-like element, but not the native terminator and the gfp reporter gene (Fig. 5A). The plasmid pF2-gfp containing the terminator-less fragment F2 drives GFP-expression in a large part of the bacterial population at both BrrG native expression and overproduction conditions (Figs. 3B and 5B). In contrast, the derivative pF2-TBBa_1006-gfp containing in addition the artificial terminator showed a strongly reduced population expression of GFP, but a marked increase in GFP-expressing bacteria upon BrrG overproduction (Fig. 5B, right panel, Suppl. Figure 6). Together, these findings demonstrate that the presence of BrrG allows the RNAP to bypass both the native termination site, as well as an unrelated synthetic terminator, indicating that BrrG is indeed a processive antiterminator.

Antiterminator BrrG allows RNA polymerase to bypass an engineered termination site.

(A) Schematic representation of the experimental setup. Plasmid pF2-TBBa_1006-gfp carries the insertion of the artificial strong terminator TBBa_1006 in between the terminator-less promoter fragment F2 and gfp of plasmid pF2-gfp. (B) Histograms of fluorescence distribution for the bacterial populations containing the fragment F2, the fragment F2 with the terminator and the sequence with the mutated -10-like element (pF6- 10mut). Time of the measurements is indicated. Induction was done with 25 ng/ml aTc. Bacteria were cultivated in M199 medium supplemented with 10% FCS and aTc inducer at indicated concentrations. Experiments were performed in biological triplicates.

Model of BrrG-antitermination.

(A) Schematic representation of terminator-dependent transcription inhibition. RNAP initiates transcription at the TSS (+1) and proceeds until the termination site (TbgtA-K), where it falls off. (B) Transcriptional antitermination of the BaGTA in presence of BrrG. BrrG acts as antiterminator enabling the transcription of the downstream bgtA-K cassette.

Discussion

Processive antitermination is an important mechanism of transcriptional control identified both for bacteria and phages. In temperate phages processive antiterminators, i.e. phage lambda N and Q-antiterminators, switch on expression of early and late genes, respectively, and thus enable the progression of the infection cycle (Goodson and Winkler, 2018).

Given the ancestral relationship between GTAs and phages, this study investigates whether transcription attenuation and antitermination are preserved in the chimeric BaGTA and whether they play a role in the regulatory integration of two functional gene clusters of divergent phage origin. We showed that the “late” capsid producing bgtA-K genes belong to a single transcriptional unit derived from one phage, and that their expression is enabled by the antiterminator BrrG, encoded within the ror locus, which originates from a second phage. However, the question remains: Is BrrG a Q-like antiterminator, and does it function in a similar manner?

The canonical Q-dependent regulatory circuit consists of several critical components: Q antiterminator, the promoter PR’, a termination site and a -10-like element (Perdue and Roberts, 2011; reviewed in Casjens and Hendrix, 2015). The first component is the Q antiterminator which binds to DNA and engages into the complex with RNAP. The Q-RNAP complex is termination-resistant and permits RNAP to bypass the termination signals in front of late genes which results in the synthesis of the capsid components (Guo et al., 1991; Grayhack and Roberts, 1982). In our study we showed that BrrG has structural similarity to the Q21 antiterminator of the P21 lambdoid phage. Moreover, BrrG is able to counteract transcription attenuation imposed by an introduced artificial termination site. Additionally, we demonstrated that BrrG binds in proximity to the PbgtA-K promoter and upregulates transcription of the bgtA-K cluster – BaGTA’s counterpart of the late gene cluster.

The second component of the canonical Q-dependent regulatory circuit is the promoter PR’ containing Q-binding site for interaction with Q and engagement with RNAP. In this study, we demonstrated that the PbgtA-K promoter overlaps with the BrrG binding site. Given that the promoter PbgtA-K is located 864 bp upstream from the first gene of the bgtA-K cassette and is separated from the former by a gene BH14100, this raises the question of whether BH14100 belongs to the bgtA-K cluster and what functional role it may play. Previous studies have shown that BH14100 is not conserved among Bartonellae and thus is likely not involved in the BaGTA regulation or synthesis (Guy et al., 2013; Tamarit et al., 2018). In support of this, our previous study using multiplexed transposon sequencing showed no effect of BH14100 disruption on GTA transfer, indicating that BH14100 has either no or only a very modest effect on GTA synthesis (Québatte et al., 2017).

The third component of the canonical Q-dependent regulatory circuit is the termination site which attenuates expression of the late gene cluster’. We showed that the putative terminator TbgtA-Kis located downstream of the PbgtA-K promoter and regulates expression of the bgtA-K operon. Whereas the bgtA-K locus is only active in the GTA producing cells, deleting the putative terminator TbgtA-Krenders broad and constitutive expression of the bgtA-K locus which suggests that the temporal regulation and heterogeneity in expression are granted by TbgtA-K.

The fourth component of the canonical Q-dependent regulatory circuit is the -10-like element located in proximity to the promoter PR’ driving expression of late genes. This element resembles the -10-element of the promoter and defines the sigma-dependent pausing of RNAP which is crucial for the engagement of Q into the complex with RNAP (Perdue and Roberts, 2011). We found that expression of the bgtA-K locus is dependent on the conserved -10-like element located downstream of the PbgtA-K promoter. Altogether, these data suggest that BrrG is an antiterminator found in BaGTA and involved in regulating expression of the bgtA-K cluster. Based on these lines of evidence we propose BrrG as Q-like processive antiterminator.

However, why does GTA-mediated DNA transfer remain unaffected upon brrG overexpression?

A simple explanation could be that bacterial lysis and particle release are controlled independently of BrrG which activates only transcription of the bgtA-K locus. This notion is supported by the apparent absence of lysis of fluorescent bacteria upon BrrG overproduction (Fig. 1D). For the archetypical RcGTA of Rhodobacter capsulatus, particle synthesis and release are controlled differentially by the CckA-ChpT-CtrA phosphorelay system, where the phosphorylation state of CtrA controls the progression from the GTA synthesis to a lytic release (Farrera-Calderon et al., 2021; Fogg, 2019; Westbye et al., 2018; Westbye et al., 2013). We hypothesize that similarly to RcGTA a trigger could be involved in the regulation of bacterial lysis and BaGTA release. BaGTA contains two putative lysis cassettes – bgtAB and bgtTUV – and only the former one is located within the larger bgt cluster, suggesting that the latter cassette is controlled independently.

Taken together, we demonstrate that the transcriptional control mechanism involving transcription attenuation and antitermination, as described for phage lambda, is functionally conserved in the chimeric BaGTA. This mechanism specifically coordinates the temporal expression of the ROR and bgtA-K loci, which have independent phage origins. The ROR-encoded BrrG acts as a processive antiterminator, facilitating the induction of the bgtA-K transcriptional unit. However, is this regulatory mechanism also conserved in other types of GTA beyond BaGTA of Bartonella? The recent publication by Tran and Le has shown that termination-antitermination is involved in control of gene expression in evolutionary distinct CcGTA of Caulobacter crescentus (Tran and Le, 2024). The authors showed that transcription of the structural genes, which are counterparts of the bgtA-K locus, is co-activated by GafY and IHF, while the GafZ antiterminator prevents expression attenuation at the termination sites. Furthermore, GafYZ regulators of CcGTA are homologous to N’- and C’-domains of the highly conserved GafA – the direct activator of RcGTA (Fogg, 2019; Gozzi et al., 2022; Sherlock and Fogg, 2022), which indicates that RcGTA might also be controlled by processive antitermination. Previous studies of GafA indicated that its C’-domain has similarity to the region 4 of a sigma factor (R4σ) and therefore authors proposed GafA as an alternative sigma factor or a transcription factor (Fogg, 2019; Sherlock and Fogg, 2022). However, the related R4σ-like fold was also found in the resolved structure of Q21 antiterminator indicating that findings of Fogg and Sherlock might, in fact, speak for the implication of GafA in antitermination (Shi et al., 2019). Interestingly, Tran and Le found neither sequence nor structural similarity of GafZ to known antiterminators Qλ and Q21. Our comparison of resolved structure of Q21 and modelled BrrG, GafZ and GafA, clearly indicates similar structural features, suggesting both GafZ and GafA being antiterminators (Suppl. Figure 1).

RcGTA and CcGTA, on one hand, and BaGTA on the other, belong to two evolutionary distinct groups of GTA that emerged from independent events of phage domestication (Kogay et al., 2022; Tamarit et al., 2018). The reliance of these groups on the termination-antitermination mechanism of transcription control clearly suggests that processive antitermination is a common regulatory mechanism in GTAs, which evolved independently in two types of GTA within Alphaproteobacteria. To date, five types of GTAs have been identified in both bacteria and archaea. Given their phage heritage, it remains unclear whether other GTA types also depend on processive antitermination for regulation, and this should be explored in future studies.

Materials and Methods

Bacterial Strains and growth conditions

All bacterial strains, plasmids and oligonucleotides used in this study are listed in Supplementary Tables 1, 2 and 3, respectively. B. henselae strains were grown at 35°C on Columbia blood agar (CBA) supplemented with 5% sheep blood, appropriate antibiotics (gentamicin (Gm) – 10 μg/ml, kanamycin (Km) – 30 μg/ml, streptomycin (Sm) – 100 μg/ml, spectinomycin (Sp) – 50 μg/ml) and other supplements (isopropyl β-D-thiogalactosidase (IPTG) – 500 μM, anhydrotetracycline (aTc) – 10 ng/ml or 25 ng/ml) in a humidified atmosphere containing 5% CO2. B. henselae strains were stored as cryostocks at -80°C and prior the experiment seeded on CBA for 3 days and subsequently expanded and grown on fresh plates for 2 days. E. coli strains were stored as cryostocks at -80°C and were grown at 37°C overnight in lysogeny broth (LB) or on solid LB agar plates (LA) supplemented with appropriate antibiotics (ampicillin (Ap) – 100 μg/ml, Gm – 10 μg/ml, Km – 30 μg/ml) and other supplements (2,6-diaminopimelic acid (DAP) – 1 mM, 1% glucose).

Construction of strains

All bacterial strains used in this study are listed in Table 1.

DNA manipulations were performed according to standard techniques and all cloned inserts were DNA sequenced to confirm sequence integrity. Chromosomal insertions or deletions in B. henselae were generated by a two-step gene replacement procedure as described in Schulein and Dehio (Schulein and Dehio, 2002). Plasmids were introduced into B. henselae strains by conjugation with E. coli strain JKE201 (MFDpir) using two-parental mating as described previously (Harms et al., 2017).

Construction of plasmids

All plasmids used in this study are listed in Table 2 and all oligonucleotides used in this study are listed in Table 3. Plasmids generated in this study were constructed as follows:

pQN011: Plasmid pTR1000 was digested with the restriction enzyme BamHI. Homologous fragments 1 (upstream) and 2 (downstream) were PCR amplified from heat-killed B. henselae cells with primers prQN029 and prQN030, prQN031 and prQN032, respectively.

Those two fragments were cloned together by performing overlap extension PCR with primers prQN030 and prQN032. The fragment and the backbone were fused with the 5X In-Fusion® HD Enzyme Premix kit.

pQN018: brrG was PCR amplified from heat-killed B. henselae cells with primers prQN053 and prQN054. tetR was PCR amplified with primer prAK227 and prAK208 from 705-bp tetR-Ptet promoter template ordered as a synthetic fragment from Genescript. The PCR product generated by amplifying pBZ485_a_empty with prAK089 and prAK109 was used as a backbone. Three fragments were combined with the 5X In-Fusion® HD Enzyme Premix kit.

pAK126: brrG was PCR amplified from heat-killed B. henselae cells with primers prAK440 and prAK441. tetR was PCR amplified with primer prAK444 and prAK208 from 705-bp tetR-Ptet promoter template ordered as a synthetic fragment from Genescript. The PCR product generated by amplification of pPG100 with prAK116 and prAK167 was used as a backbone. Three fragments were combined with the 5X In-Fusion® HD Enzyme Premix kit.

pQN003: Fragment Pf3 was PCR amplified from heat-killed B. henselae cells with primers prQN005 and prQN014. dsRed was PCR amplified from pJC44 with primer prAK097 and prAK110. The PCR product generated by amplifying pBZ485_a_empty with prAK089 and prAK109 was used as a backbone. Three fragments were combined with the 5X In-Fusion® HD Enzyme Premix kit.

pQN006: gfpmut2 was PCR amplified from plasmid pCD353 with primers prQN015 and prQN016. The PCR product generated by amplifying pBZ485_a_empty with prAK089 and prAK109 was used as a backbone. Two fragments were combined with the 5X In-Fusion® HD Enzyme Premix kit.

pQN008: Plasmid pQN006 was digested with BamHI. Fragment sF1 was PCR amplified form heat-killed B. henselae cells with primers prQN021 and prQN022. The fragment and the backbone were ligated with the 5X In-Fusion® HD Enzyme Premix kit.

pQN009: Plasmid pQN006 was digested with BamHI. Fragment sF2 was PCR amplified form heat-killed B. henselae cells with primers prQN018 and prQN017. The fragment and the backbone were ligated with the 5X In-Fusion® HD Enzyme Premix kit.

pQN010: Plasmid pQN006 was digested with BamHI. Fragment sF3 was PCR amplified form heat-killed B. henselae cells with primers prQN019 and prQN020. The fragment and the backbone were ligated with the 5X In-Fusion® HD Enzyme Premix kit.

pQN020: The minimal predicted PGTA promoter was introduced into plasmid pQN006 by PCR with the primers prQN061 and prQN062. The plasmid was circularized with the 5X In-Fusion® HD Enzyme Premix kit.

pQN057: The PCR product generated by amplifying pQN006 with prQN111 and prAK109 was used as a backbone. F1 was PCR amplified from heat-killed B. henselae cells with primers prQ134 and prQN135. F1 and the backbone were combined with the 5X In-Fusion® HD Enzyme Premix kit.

pQN058: The PCR product generated by amplifying pQN006 with prQN111 and prAK109 was used as a backbone. F2 was PCR amplified from heat-killed B. henselae cells with primers prQ134 and prQN136. F2 and the backbone were ligated with the 5X In-Fusion® HD Enzyme Premix kit.

pQN059: The PCR product generated by amplifying pQN006 with prQN111 and prAK109 was used as a backbone. F3 was PCR amplified from heat-killed B. henselae cells with primers prQ134 and prQN137. F3 and the backbone were ligated with the 5X In-Fusion® HD Enzyme Premix kit.

pQN060: The PCR product generated by amplifying pQN006 with prQN111 and prAK109 was used as a backbone. F4 was PCR amplified from heat-killed B. henselae cells with primers prQ134 and prQN138. F4 and the backbone were ligated with the 5X In-Fusion® HD Enzyme Premix kit.

pQN061: The PCR product generated by amplifying pQN006 with prQN111 and prAK109 was used as a backbone. F5 was PCR amplified from heat-killed B. henselae cells with primers prQ134 and prQN139. F5 and the backbone were ligated with the 5X In-Fusion® HD Enzyme Premix kit.

pQN007: Fragment F6 was PCR amplified from heat-killed B. henselae cells with primers prQN005 and prQN014. Plasmid pQN006 and F6 were restricted with BamHI and Xmal and ligated with RAPID DNA Dephos and Ligation Kit.

pQN047: The mutated -10-like sequence was introduced into plasmid pQN007 by PCR with primers prQN120 and prQN121. The plasmid was circularized with the 5X In-Fusion® HD Enzyme Premix kit.

pQN048: N’-3x-FLAG was introduced into pQN018 by PCR with primers prQN114 and prQN115. The plasmid was circularized with the 5X In-Fusion® HD Enzyme Premix kit.

pQN049: C’-3x-FLAG was introduced into pQN018 by PCR with primers prQN116 and prQN117. The plasmid was circularized with the 5X In-Fusion® HD Enzyme Premix kit.

pAK180: Terminator TBBa_1006 was introduced into pQN058 by PCR with primers prAK592 and prAK593. The linear PCR product was used for the direct transformation of E. coli.

Measuring expression of the BaGTA locus with flow cytometry

B. henselae strains were grown as described above. Bacteria were resuspended in M199 supplemented with 10% FCS and other additives, e.g. inducer, if required (hereafter M199+), diluted to OD600 nm = 0.008 and incubated in 24-well plate for a defined amount of time in a humidified atmosphere at 35°C and 5% CO2. After a defined amount of time bacteria were collected, centrifuged, and washed once with PBS. Bacteria were resuspended in 4% PFA and fixed for 7 min at RT. The fixation was stopped by adding PBS supplemented with 2% FCS. Samples were kept at 4°C in the dark before flow cytometry analysis. GFP fluorescence in samples was recorded with the BD LSRFortessa™ Cell Analyzer at the FACS Core Facility, Biozentrum, the University of Basel. Data were analysed using software FlowJo (Version 10.8.1). Visualisation of the data was done with GraphPad prism 8.

GTA transfer assay

GTA transfer assays were done as co-cultivations of two B. henselae strains. B. henselae strains were grown as described above. Bacteria were resuspended in M199+ supplemented, if required, with inducer at specified concentrations and diluted to OD600 nm = 0.008. For tested co-cultivation pair 1 ml of each strain was mixed and a mixture was incubated in 24-well plate for 24 h in a humidified atmosphere at 35°C and 5% CO2. After 24 h bacteria were collected and centrifuged. Bacterial pellets were resuspended in 30 μl of M199+ and resuspensions were used for serial dilutions followed by plating onto CBA plates supplemented with SmGm or SmGmSp antibiotics. Bacteria were grown for 6-15 d in a humidified atmosphere at 35°C and 5% CO2. Bacterial colonies were counted, and CFU/ml was determined for each co-cultivation pair. Frequency of double resistant bacteria (DRB) was calculated as the ratio of CFUs of recipient DRB (SmR GmR SpR) / total number of recipients (SmR GmR), yielding the transfer frequency per recipient cell.

Chromatin immunoprecipitation coupled with high throughput sequencing

B. henselae strains were grown as described above. Bacteria were harvested from plates and resuspended in 5 ml of Schneider’s Insect Medium (24.5 g/l Schneider’s Drosophila Powder Medium Revised, 5% (wt/vol) sucrose, 0.025 M HEPES, adjusted pH 7.25) supplemented with 10% FCS (hereafter Schneider’s Medium) (Guy et al., 2013; Riess et al., 2008). Bacteria were diluted to OD600 nm = 0.01 in 500 ml Schneider’s Medium for each strain and grown at 35°C with shaking 135 rpm for 4 d until OD600 nm = 0.6-0.7. Bacteria were pelleted and pellets were resuspended in 50 ml PBS. Bacteria were fixed with PFA with final concentration 1% and incubated for 10 min at RT. The fixation was stopped by adding glycine to a final concentration of 125 mM, and mixture was incubated for 20 min on ice. Fixed bacteria were centrifuged for 30 min at 4°C at 2000 x g, and bacterial pellets were frozen at -80°C. Thawed pellets were resuspended in 10 ml of ChIP Buffer (1.1% Triton X-100, 1.2 mM EDTA, 16.7 mM Tris-HCl (pH 8.1), 167 mM NaCl, supplemented with cOmplete™Mini EDTA-free Protease Inhibitor Cocktail) and cells were lysed by 4 passages with French Press. The lysate was sonicated, and DNA size was checked with the 1.5% agarose gel. The lysate was centrifuged with the maximal speed and supernatant was collected. A 50 μl aliquot of supernatant was taken as “Input DNA” for the subsequent sequencing and data analysis. Following manufacturers’ protocols, remaining supernatant was immunoprecipitated ON with anti-FLAG Magnetic Agarose or anti-HA magnetic beads for 3xFLAG-tagged and HA-tagged proteins, respectively. Magnetic beads/agarose were immobilized and after removal of supernatant were sequentially rinsed with a panel of buffers: Low Salt Wash Buffer (0.1% SDS, 1% Triton X-100, 2 mM EDTA, 20 mM Tris-HCl (pH 8.1), 150 mM NaCl), High Salt Wash Buffer (0.1% SDS, 1% Triton X-100, 2 mM EDTA, 20 mM Tris-HCl (pH 8.1), 500 mM NaCl), LiCl Wash Buffer (250 mM LiCl, 1% NP-40, 1% deoxycholate, 1 mM EDTA, 10 mM Tris-HCl (pH 8.1)), TBS Buffer (50 mM Tris-HCl, 150 mM NaCl). The thawed Input control and immobilized Magnetic beads/agarose were mixed with 450 μl of TBS buffer, 15 µl of 5 M NaCl and 0.5 μl RNAse A (100mg/ml). The mixtures were incubated at 65°C ON. Then 4 µl of 0.5 M EDTA and 2 µl of 10 mg/ml Proteinase-K were added and the mixtures were incubated at 45°C for 2 h. The resulting mixtures were thoroughly mixed in 1:1 ratio with phenol/chlorophorm/isoamylalcohol combination and centrifuged at >10 000 x g for 1 min. The upper aqueous phase was transferred to a new tube, mixed in 1:1 ratio with chloroform and centrifuged at >10’000 x g for 1 min. The upper aqueous phase was transferred to a new tube and mixed with: (i) 25 μg of linear polyacrylamide; (ii) NaAc to a final concentration of 0.3 M; (iii) in 1:1 ratio with 100% isopropanol. The solution was centrifuged at >10’000 x g for 30 min, washed twice with cold 70% EtOH. Pelleted DNA was dried and resuspended in water ON. Library preparation and sequencing were done at the Genomics Facility Basel, D-BSSE, ETH-Zurich.

Bioinformatics analysis

All alignments were done with the Clustal Omega and MUSCLE implemented in Geneious Prime 2020.2.3 (Edgar, 2004; Sievers et al., 2011). Mapping of ChIP-Seq data was done with bowtie2 (Langmead and Salzberg, 2012). Converting SAM to BAM, extracting unique binding reads and sorting were done with SAMtools (Danecek et al., 2021). Coverage plots were generated with bamCoverage and visualized with the Integrative Genomics Viewer (Ramírez et al., 2016; Robinson et al., 2011). Promoter predictions were done with BProm (Solovyev et al., 2011). Models of protein structures were either done with AlphaFold2 or taken from AlphaFold Protein Structure Database (Jumper et al., 2021; Varadi et al., 2022). Comparing the predicted models to Protein Data Bank (PDB) was done with the Dali server (Holm, 2022). Predictions of the secondary structure of the putative terminator was done with Mfold with the default parameters (Zuker, 2003). Calculations were performed at sciCORE (http://scicore.unibas.ch/) scientific computing centre at the University of Basel.